% 02-abbreviations.tex - Section 2: Abbreviations

\chapter{Abbreviations and Definitions}
\label{sec:abbreviations}

\section{Abbreviations}

\begin{longtable}{p{3cm}p{10cm}}
    \toprule
    \textbf{Abbreviation} & \textbf{Definition} \\
    \midrule
    \endfirsthead

    \toprule
    \textbf{Abbreviation} & \textbf{Definition} \\
    \midrule
    \endhead

    \bottomrule
    \endfoot

    ADC & Analog-to-Digital Converter \\
    ASP & Analog Signal Processor \\
    BAM & Beamformer coefficient command \\
    COR & Correlator command/mode \\
    DRX & Digital Receiver command/mode \\
    FEE & Front End Electronics \\
    FFT & Fast Fourier Transform \\
    FPGA & Field-Programmable Gate Array \\
    GPU & Graphics Processing Unit \\
    ICD & Interface Control Document \\
    IPMI & Intelligent Platform Management Interface \\
    LWA & Long Wavelength Array \\
    M\&C & Monitor and Control \\
    MCS & Monitor and Control System \\
    MIB & Management Information Base \\
    MJD & Modified Julian Date \\
    MPM & Milliseconds Past Midnight \\
    NDP & Next-Generation Digital Processor \\
    PFB & Polyphase Filter Bank \\
    TBS & Transient Buffer -- Streaming \\
    TBT & Transient Buffer -- Triggered \\
    TCC & Tensor Core Correlator \\
    UDP & User Datagram Protocol \\
    ZCU & Zynq UltraScale+ MPSoC (ZCU102 board) \\

\end{longtable}

\section{Definitions}

\begin{description}
    \item[Stand] A single antenna structure consisting of two polarizations (X and Y). Each stand is identified by a unique number from 1 to 256.

    \item[Input] A single signal path from one polarization of one stand. Inputs are numbered 1 to 512, where input $n$ corresponds to stand $\lceil n/2 \rceil$ and polarization $(n-1) \mod 2 + 1$.

    \item[Board] A ZCU102 FPGA board that services 16 stands (32 inputs). Boards are numbered 1 to 16.

    \item[Server] A GPU-equipped processing server running the X-engine software. Servers are numbered 1 to 4.

    \item[Head node] A GPU-equipped processing server running the NDP control and T-engine software.

    \item[Pipeline] A processing instance within a server. Each server runs 4 pipelines.

    \item[Beam] A phased-array beam output. Beams are numbered 1 to 4, each with two polarizations.

    \item[Tuning] A frequency setting for a beam. Each beam supports up to 2 tunings.

    \item[Channel] A frequency channel produced by the PFB. Channels are numbered 0 to 4095, with channel $n$ centered at frequency $n \times f_c$ Hz, where $f_c \approx 23.926$~kHz (196000 kHz / 8192).

    \item[Subslot] A time subdivision of a second. There are 100 subslots per second (10~ms each).

    \item[Slot] A one-second time interval used for command scheduling.

\end{description}

\section{Fixed Parameters}

\begin{siteNote}
The parameters shown below are for a full station like LWA1.  For a mini-station the values for NSTAND and NINPUT will be one-forth of the presented values.  There may also be other differences in NPIPE_PER_SERVER.
\end{siteNote}

Table~\ref{tab:fixed-params} lists the fixed parameters used throughout this document.

\begin{table}[htbp]
    \centering
    \caption{NDP Fixed Parameters}
    \label{tab:fixed-params}
    \begin{tabular}{lll}
        \toprule
        \textbf{Symbol} & \textbf{Value} & \textbf{Description} \\
        \midrule
        SUBSYSTEM & ``NDP'' & Subsystem identifier \\
        \fs{} & 196.0~MHz & Signal sampling frequency \\
        \fc{} & 23.926~kHz & Channel bandwidth ($f_s / (2 \times 4096)$) \\
        NCHAN & 4096 & Number of frequency channels \\
        NSTAND & 256 & Number of stands \\
        NPOL & 2 & Polarizations per stand \\
        NINPUT & 512 & Total number of inputs \\
        NBEAM & 4 & Maximum number of beams \\
        NBOARD & 16 & Number of F-engine boards \\
        NSERVER & 4 & Number of X-engine servers \\
        NPIPE\_PER\_SERVER & 4 & Pipelines per server \\
        ADC\_BITS & 8 & ADC resolution \\
        \bottomrule
    \end{tabular}
\end{table}
